\chapter{Conclusion}
\label{chapter:conclusion}

\section{Summary}

Modern computer architectures execute concurrent programs on shared memory, and the subtle behaviors that arise under weak memory models can lead to hard-to-find bugs and security vulnerabilities. The Cat specification language provides a concise axiomatic framework for describing these models as graphs of binary relations, and many architectures --- including x86, RISC-V, Arm, and the Linux kernel --- already have Cat definitions. However, mechanizing these specifications in a proof assistant and constructing formal proofs about them have remained largely orthogonal problems. Writing proofs from scratch for each architecture is time-consuming, and most existing mechanizations do not reuse Cat files directly, introducing additional translation work.

In this thesis, we designed and implemented LeanCat, a domain-specific proof framework that addresses both problems simultaneously. LeanCat parses existing Cat specification files and introduces their semantics into Lean 4 via its metaprogramming framework, so users do not need to translate or rewrite memory model definitions by hand. The \texttt{defcat} command elaborates a Cat file into native Lean 4 definitions, and the resulting predicates integrate directly with Lean 4's type system and tactic infrastructure.

We defined the core types needed to represent candidate executions in Lean 4 --- events with optional access tags, candidate execution records, and a suite of well-formedness conditions --- and we developed a library of theorems about binary relations and acyclicity that bridge Lean 4's existing Mathlib library and the specific needs of weak memory model reasoning. The central result in this library is \texttt{acyclicMono}: if a relation is acyclic and a second relation is a subset of it, then the second relation is also acyclic. This monotonicity lemma is the key proof technique for comparing memory models, and it reduces what would otherwise be a complex semantic argument to a straightforward subset check on relation definitions.

We demonstrated LeanCat through two families of case studies. First, we checked concrete litmus test executions against several memory models. For SC, TSO, MIPS, and LKMM, we constructed candidate executions for the Store Buffering, Message Passing, Load Buffering, and CoRR tests and proved, using \texttt{acyclic\_of\_rank} and related lemmas, whether each outcome is allowed or forbidden. These case studies confirm that our formalization captures the intended semantics: SB separates SC from TSO and MIPS, MP separates TSO from MIPS, and LB is forbidden by all three. Second, we formally compared memory models with each other. In particular, we proved the theorem \texttt{scvtso}, a machine-checked proof that every execution allowed by SC is also allowed by TSO, together with a counterexample showing that the inclusion is strict. This gives a complete account of the SC--TSO relationship: SC is strictly stronger than TSO, and the proof reduces entirely to a monotonicity argument on the acyclicity constraints of the two models.

We also integrated execution graph visualization into the Lean 4 development environment using ProofWidgets, allowing users to inspect candidate executions graphically during proof development. This makes it significantly easier to identify cycles by eye before committing to a formal proof.

Taken together, these results show that LeanCat achieves its stated goal. It significantly reduces the mechanization effort and boilerplate required for weak memory model proofs, and the framework generalizes naturally: the same infrastructure --- \texttt{defcat}, \texttt{acyclicMono}, and the candidate execution types --- applies uniformly across SC, TSO, MIPS, and LKMM without architecture-specific proof engineering.

\section{Future Work}
One promising direction for future work is to extend LeanCat to support more complex model comparisons, such as between RISC-V and TSO, or between LKMM and the underlying architectures it is designed to support. As we discussed in Section~\ref{sec:limitations}, the Cat language leaves many properties implicit because herdtools7 handles them computationally. For example, Cat assumes coherence order is a strict total order per location without stating this formally. In Lean 4, every such assumption must be made explicit as a wellformedness condition before proofs become possible. For simple models like SC and TSO, the number of such implicit assumptions is manageable. However, for more complex architectures like RISC-V, the number of implicit assumptions grows significantly, and identifying and formalizing them correctly requires substantial additional effort. LeanCat already provides the foundational infrastructure for such comparisons --- the \texttt{defcat} command, the \texttt{acyclicMono} lemma, and the candidate execution framework are all in place. The remaining work is to identify the implicit assumptions for each new model and add them as wellformedness conditions, which is a well-defined engineering task rather than a fundamental research challenge. We therefore expect that extending LeanCat to support RISC-V and other architectures is achievable in future work, building directly on the infrastructure developed in this thesis.