\chapter{Implementation in Lean 4}
\label{chapter:design}

When we work with pen and paper, we usually do not focus on the subtle parts of a proof; instead, we prioritize the interesting aspects. In contrast, mechanization requires us to define every detail of the specification. These challenges are independent of the particular proof assistant we choose, so we propose a general design whose ideas can be easily implemented in other languages. In this section, we identify and clarify the issues involved in mechanizing weak memory models and show how we address them using Lean 4’s existing MathLib and Meta-Programming framework [\citet{deMoura2021lean4,10.1145/3110278Lean4Meta,mathlib2020}].

\begin{figure}[t]
  \centering
  \begin{minipage}{0.95\linewidth}
\begin{verbatim}
include "cos.cat"
let com = rf | fr | co
acyclic po | com as sc
\end{verbatim}
  \end{minipage}
  \caption{Cat definition of sequential consistency (\textsc{sc}).}
  \label{fig:cat-sc}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{minipage}{0.95\linewidth}
\begin{verbatim}
include "cos.cat"
let com-tso = rf | co | fr
let po1-tso = po & (W*W | R*M)
let ghb = po1-tso | com-tso
acyclic ghb as tso
\end{verbatim}
  \end{minipage}
  \caption{Cat definition of total store order (\textsc{tso}).}
  \label{fig:cat-tso}
\end{figure}

\section{Core Types for Memory Models in Lean 4}
\label{sec:design:core-types}
To parse the weak memory models, we need to define the data types we operate on firstly. We use \citet{catOnlineDocHerd,catManualSyntax11} as our reference. The \citet{catManualSyntax11} defines the syntax and semantics of the Cat language in a formal way, but it is not always up to date. For example, it does not define the \texttt{[]} unary operator, while the \citet{catOnlineDocHerd} defines it and is continuously updated. Therefore, most of the time we use the online documentation as our primary reference.

\subsection{Event and Set}
Events and sets of events are the most fundamental data types in the Cat language: Cat defines predicates and binary relations over them in \citet{catOnlineDocHerd}. One event represents an instruction, so it has an ID, a thread ID, an address, a value, etc. We can represent the thread ID and the ID using \texttt{Nat}, but the address and value are not fixed by Cat. Different architectures may have different address spaces; for example, some machines use 32-bit addresses, while others use 64-bit addresses or even wider representations. Similarly, values may range over signed integers or floating-point numbers. These details are intentionally left unspecified by Cat \citet{catOnlineDocHerd,catManualSyntax11}.

However, addresses and values are irrelevant to the problem we want to solve. LeanCat works at a high level of abstraction: we do not need to generate a DSL or assembly code to simulate litmus tests; instead, we only care about the execution graph and the relations between instructions in the program. Therefore, in our design we use \texttt{Nat} to represent both addresses and values.

In cat, we use Set in Lean 4 to organize the events, and there are some primitive set of events like W (All the write events), R (All the read events), F (All the fence events), etc. These events are gathered from the herd7 tools for each litmus tests, and they call them anarchical semantics, we can simply treat them as the input of the model. Besides the primitive events, there are some sets are composed by these primitive sets, for example, the M (Set of modification events) is the union of the W and R. Besides that, there are some user-defined sets, which are an important part of the LKMM model; these sets are defined by the tag carried by each event.

\subsection{Tagged Events}

There are many sets; they are defined by tags instead of being implicitly defined in the Cat language. For example, in LKMM memory models, there are relations used to express accesses; they are normally defined in this way:

\begin{verbatim}
enum Accesses = 'ONCE (*READ_ONCE,WRITE_ONCE*) ||
                'RELEASE (*smp_store_release*) ||
                'ACQUIRE (*smp_load_acquire*) ||
                'NORETURN (* R of non-returning RMW *) ||
                'MB (*xchg(),cmpxchg(),...*)
instructions R[Accesses]
instructions W[Accesses]
instructions RMW[Accesses]
\end{verbatim}

The \texttt{Accesses} are then used to tag the events that are in \texttt{R} (reads), \texttt{W} (writes), or \texttt{RMW} (read-modify-writes). These events serve an important role, because they can be used to disambiguate the types of operations in each case. For example, we can identify a read as an acquire-read by tagging a read event with \texttt{ACQUIRE}. Because weak memory model semantics represent only executions, in that case only one tag can be attached. Also, even if an \texttt{R} event was produced by a regular read instruction or an acquire, this is why we need a tag system.

We use the Sigma type to represent the tag system in this way:
\begin{verbatim}
structure Event where
  id : Nat   -- Unique identifier
  t_id : Nat -- Thread ID
  effect : Effect -- Action performed
  tag : Sigma tagType : Type, tagType
\end{verbatim}

The Sigma type is a dependent type in which the first component is the type on which the second component depends. Since we only know the tag when we know the explicit candidate executions, we cannot determine the types of the tags before parsing the Cat files, as the tags are normally defined in the Bell files. We only know the explicit type when we see the definitions, and the Sigma type captures this property well, because the actual tag used in the event is determined by the tag type defined in Bell files.

However, the Sigma type introduces another problem: \texttt{Event} cannot derive \texttt{DecidableEq}. \texttt{DecidableEq} is used frequently in practice to automatically determine whether two instances are equal. Normally this can be derived directly, but the problem with this event definition is the Sigma type: it introduces a type as an argument, so \texttt{Event} cannot derive \texttt{DecidableEq} automatically. This is because all the types in \texttt{Event} must also have \texttt{DecidableEq}, and there is no constraint that \texttt{tagType} has \texttt{DecidableEq}.

To solve this, we use an axiom:
\begin{verbatim}
axiom event_id_unique :
  ∀ e₁ e₂ : Event, e₁.id = e₂.id -> e₁ = e₂
\end{verbatim}

It states that if two events have the same \texttt{id}, then they are the same event. This is a little bit unsafe, because it is easy to define two different events that have different operations but the same \texttt{id}. However, this is one of our assumptions: we assume all events provided by litmus tests are distinguishable by their \texttt{id}s. With this axiom, we can define \texttt{DecidableEq} for \texttt{Event}. It is a trade-off between practical usability and soundness.

Because the macro system is defined as a translation from one syntax to another, each tag definition is compiled into a corresponding \texttt{def} in Lean 4. However, this design exposes a limitation: Lean 4 does not support global renaming of definitions. This becomes problematic because the translation is performed on each instruction independently.

In particular, when multiple instructions share the same tag set but require different constraints—such as \texttt{F[Accesses]}, \texttt{W[Accesses]}, and \texttt{RMW[Accesses]}—we would ideally refine or rename \texttt{Accesses} per instruction to reflect these constraints. Since Lean 4 does not permit such global renaming, this approach is not feasible.

To address this issue, we redesign the grammar. Instead of assigning tags to a single instruction using the form \{\textit{event}\}[\textit{tag}], we generalize it to \{\textit{events}\}[\textit{tags}], where \textit{events} denotes a set of instructions. This allows us to attach a shared tag constraint to multiple instructions simultaneously, avoiding the need for renaming.

For example, the previous cases can be expressed uniformly as \{\texttt{F, W, RMW}\}[\texttt{Accesses}]. Under this formulation, the \texttt{ACQUIRE} case can be represented as:

\begin{verbatim}
fun evts x a => 
  a ∈ {e | e.tag = ⟨lkmm.Accesses, lkmm.Accesses.ACQUIRE'⟩} ∩ x.evts.R ∩ x.evts.W 
  ∧ a ∈ x.evts.RMW
\end{verbatim}

\subsection{Candidate Execution}
Candidate executions play a crucial role in the Cat language: they represent a specific possible execution and may comprise an execution graph. \citet{catManualSyntax11} abstracted each execution as a candidate execution, providing sets of events, each of which satisfies a specific property, such as read-from, initial writes, and so on. We separately represent these sets: a candidate execution contains a set of events as a parameter, and it also includes primitive relations that can be used to filter these events.

In addition to these definitions, we treat the coherence order and from-read relations as primitive relations, following their definitions in the Cat standard library. The standard library defines the coherence order as a series of computations, but this computation is difficult to represent in Lean~4. For example, defining the coherence order requires the use of an operation called \texttt{cross}:
\[
  \mathit{cross}(S_1 \mathbin{+\!\!+} S)
  \;=\;
  \bigcup_{\substack{e_1 \in S_1 \\ t \in \mathit{cross}(S)}}
  \left\{
    e_1 \cup t
  \right\}.
\]

In the definition above, $\mathit{cross}(S_1 \mathbin{+\!\!+} S)$ is constructed by taking the set of all unions of one relation $e_1$ chosen from $S_1$ and one relation $t$ chosen from $\mathit{cross}(S)$. This computation is difficult to represent and reason about effectively. First, it is unclear how to express it using the current definitions of sets in Lean. Second, all of its associated properties would need to be proved explicitly, while providing limited practical benefit. Therefore, for most relations of this kind in the standard library, we define them directly as primitive relations instead of computing them.

We define the candidate execution as follows:
\begin{verbatim}
structure CandidateExecution (evts : Events) where
  evts := evts
  idUnique := ∀ e₁ e₂ : Event, (e₁ ∈ evts ∧ e₂ ∈ evts) -> e₁.id ≠ e₂.id
  po   := evts.po
  rf   : SetRel Event Event
  rfInst : wellformed.rf evts rf
  co   : SetRel Event Event
  [preCo : wellformed.co evts co]
  fr   := rf.inv.comp co
  rmw  : SetRel Event Event
  wmb  : SetRel Event Event
  data : SetRel Event Event
  addr : SetRel Event Event
  ctrl : SetRel Event Event
  fence : SetRel Event Event
  mb : SetRel Event Event
  uniqueId : ∀ (e₁ e₂ : Event),
    e₁ ∈ evts.all → e₂ ∈ evts.all
    -> e₁ ≠ e₂
    → e₁.id ≠ e₂.id
\end{verbatim}

The \textit{Candidate Execution} is designed as a universal superset containing all possible relations and event types. Because different architectures, such as RISC-V \cite{riscv-spec} and Arm \cite{arm-arm} --possess distinct sets of relations and events, our framework maintains compatibility by supporting these variations within a single, unified structure. This approach is consistent with the Linux Kernel Memory Model (LKMM) \cite{alglave2018frightening}, which must account for the various behaviors of multiple underlying architectures, including dependencies such as \textit{data}, \textit{address}, and \textit{control}.

We establish a formal abstraction at this boundary: while a candidate execution is typically generated from a litmus test via a parser or compiler, the correctness of that translation is not what we focus on at the memory level. In this project, we assume the correctness of the generation process and treat the candidate execution itself as the input. It is important to note that this is a significant assumption in practice; instructions with similar behaviors may not produce identical events and relations. For example, a Read instruction might be represented as a simple $R$ in a standard \texttt{.cat} file, whereas in LKMM it is treated as \texttt{READ\_ONCE}. These differing semantics generate different candidate executions; however, by abstracting these details and focusing solely on the candidate execution as the input, we significantly reduce the overall verification effort.

\subsection{Well-Formedness Conditions}
Well-formedness conditions are the important properties that give each set and relation a meaning. Only when the relations satisfy these conditions do we say they are well-formed; otherwise they are meaningless. Most of the well-formedness conditions are adapted from the herdtools7 Coq development \cite{10.1145/2627752, 10.1145/3458926ArmedCat}.



We require the \texttt{preCo} constraint in order to introduce the coherence order relation, since it naturally satisfies the properties of a strict total order. Intuitively, the coherence order represents a sequence of write events to the same memory location. As such, it is transitive: if one write overwrites another, and that write in turn overwrites a third, then the first write necessarily overwrites the third. It is also irreflexive: since the coherence order relates two distinct instructions, it is impossible for an instruction to overwrite itself, or for two instructions to overwrite each other in both directions.



\subsection{Tactics}

Writing memory model proofs in Lean~4 involves two distinct kinds of tedium. The first is establishing the well-formedness obligations of a concrete candidate execution --- showing that the events, coherence order, reads-from relation, and so on satisfy the conditions listed in Section~\ref{sec:design:core-types}. The second is closing the individual sub-goals that arise during an acyclicity or model-comparison proof, where the goal typically decomposes into a large case split over which relation each edge belongs to. Both kinds of tedium are highly repetitive but follow predictable patterns, so they are good targets for automation.

Lean~4 provides several general-purpose tactics that are useful here. The \texttt{simp} tactic rewrites goals using a user-supplied or default set of \texttt{@[simp]} lemmas; we annotate most of our relation definitions with \texttt{@[simp]} so that \texttt{simp} can unfold them automatically. As introduced in Section~\ref{subsec:aesop}, \texttt{aesop} performs extensible proof search; in LeanCat we register \texttt{ayclicMono} with \texttt{@[aesop safe apply]} so that it is applied automatically whenever the goal is an acyclicity statement. The \texttt{decide} tactic closes goals that reduce to a decidable computation over finite data, which is useful when checking concrete event identifiers or thread identifiers. The \texttt{omega} tactic discharges arithmetic goals over the natural numbers and integers, which arises in program-order arguments because event IDs and thread IDs are encoded as \texttt{Nat}.

However, these general-purpose tactics are not sufficient on their own for the well-formedness obligations of concrete candidate executions. Establishing, for example, that the reads-from relation of the SB litmus test is well-formed requires unfolding the definition of \texttt{wellformed.rf}, inspecting each pair in the finite relation, and checking that each pair satisfies all the structural conditions (same location, write on the left, read on the right, etc.). Doing this manually for every litmus test is both tedious and error-prone, because the unfolding and case analysis follow exactly the same pattern each time.

To address this, we define a custom tactic macro \texttt{candidateExecution\_wf}:

\begin{verbatim}
macro "candidateExecution_wf" : tactic =>
  `(tactic|
    (all_goals repeat first
       | simp at *
         aesop
       | (casesm _ ∈ _
          aesop)))
\end{verbatim}

The macro repeatedly applies two strategies. The first simplifies all goals using \texttt{simp} and then attempts to close them with \texttt{aesop}. The second destructs set-membership hypotheses of the form \texttt{e ∈ S} using \texttt{casesm}, which enumerates the finitely many elements of the concrete event set, and then tries \texttt{aesop} on each case. We use it to automatically prove the well-formedness of the candidate executions.

\subsection{Models}
Now that we have the primitive types, we need a way to connect them. We can reuse existing definitions from Mathlib for working with \texttt{Set} and \texttt{Rel}, and then describe \textsc{tso} and \textsc{sc} using these functions.

The following is the standard Cat definition of sequential consistency (\textsc{sc}; see Figure~\ref{fig:cat-sc}). We ignore the \texttt{include} directive here, since we model the relations imported from \texttt{cos.cat} as primitives. The first statement, \texttt{let com = rf | fr | co}, introduces a name for the expression on the right-hand side; in Lean, this can be represented with a \texttt{def}. The operator \texttt{|} is relational union. Since Mathlib does not provide union and intersection for relations out of the box, we define them as follows:
\begin{verbatim}
def union (r1 r2 : Rel Event Event) : Rel Event Event :=
  fun x y => r1 x y \/ r2 x y

def inter (r1 r2 : Rel Event Event) : Rel Event Event :=
  fun x y => r1 x y /\ r2 x y

instance : Union (Rel Event Event) := ⟨union⟩
instance : Inter (Rel Event Event) := ⟨inter⟩
\end{verbatim}
With these instances, Lean can use the usual notation (e.g., $r_1 \cup r_2$ and $r_1 \cap r_2$) to combine relations.

The most important proposition is \emph{acyclicity}, and there are other propositions that are also useful in proofs, such as \emph{irreflexivity}. For a relation $r : \mathrm{Event} \to \mathrm{Event} \to \mathrm{Prop}$, we define acyclicity as
\[
  \mathit{Acyclic}(r)
  \;:\!\!\iff\;\;
  \forall a : \mathrm{Event},\; \neg\,\mathrm{Relation.TransGen}\; r\; a\; a.
\]
Intuitively, this means that if we follow edges of $r$, we can never return to the starting event (i.e., there is no directed cycle). This is the Cat-style formulation of acyclicity \citep{catManualSyntax11,catOnlineDocHerd}. Another useful proposition is irreflexivity,
\[
  \mathit{Irreflexive}(r) \;:\!\!\iff\; \forall a : \mathrm{Event},\; \neg\, r\; a\; a,
\]
which is already provided by Mathlib \citep{mathlib2020}.

Next, consider total store order (\textsc{tso}; see Figure~\ref{fig:cat-tso}). Compared to sequential consistency, its Cat definition uses additional operators, most notably relational intersection (written \texttt{\&} in Cat, corresponding to $\cap$) and Cartesian products of event sets (written \texttt{*} in Cat). If $S_1,S_2 \subseteq \mathrm{Event}$ are sets of events, then their product induces a relation
\[
  S_1 \times S_2
  \;:=\;
  \{\,(e_1,e_2) \mid e_1 \in S_1 \wedge e_2 \in S_2\,\}.
\]
In Lean, we implement this idea either via a definition
\begin{verbatim}
def prod (s1 s2 : Set Event) : Rel Event Event :=
  fun e1 e2 => e1 \in s1 /\ e2 \in s2
\end{verbatim}
or directly via \texttt{Set.prod}; LeanCat uses \texttt{Set.prod} to reuse existing Mathlib lemmas \citep{mathlib2020}.


\section{Connecting Cat Specifications to Lean 4}
Now that we know how to represent the types and operators of the Cat language in Lean~4, we need a way to turn this representation into a convenient DSL inside Lean. In principle, we could use the data structures defined above to encode each memory model manually and then prove properties about it. However, our goal is to reuse existing Cat specifications (e.g., Figures~\ref{fig:cat-sc} and~\ref{fig:cat-tso}), rather than rewriting them by hand \citep{catOnlineDocHerd,catManualSyntax11}.

To parse Cat files, there are several possible implementation strategies. For example, we could use an external parser (in OCaml, Rust, or C/C++) to translate Cat into an intermediate representation (such as an AST) and then generate Lean code. Instead, we rely on Lean~4's metaprogramming framework and implement the parsing pipeline inside Lean itself, which removes one translation layer and lets the kernel type-check the elaborated result \citep{10.1145/3110278Lean4Meta,deMoura2021lean4}.

More concretely, after parsing, our macros and elaborator code produce ordinary Lean definitions (e.g., \texttt{def}s of relations and consistency predicates). These generated definitions are then checked by the Lean~4 kernel just like handwritten code: if a macro expands into an ill-typed term (for example, a relation with the wrong domain/codomain, or a definition that uses an operator at an incompatible type), the kernel rejects it and the model fails to elaborate. In this sense, the kernel provides a sound defense: the trusted computing base is the small kernel, not the parsing/translation code that merely constructs candidate terms \citep{10.1145/3110278Lean4Meta,deMoura2021lean4}.

This is important for our running example (Figures~\ref{fig:cat-sc} and~\ref{fig:cat-tso}): once \textsc{sc} and \textsc{tso} are elaborated into Lean definitions, any theorem we prove about them is a proof term checked by the kernel. In particular, a statement such as \textsc{sc} $\Rightarrow$ \textsc{tso} is accepted only if the kernel verifies that the proof term has exactly the required type.

Therefore, even if our macros are imperfect, they cannot silently ``prove'' false statements: at worst they generate ill-typed terms (which are rejected), or they generate a well-typed but unintended definition (which becomes a semantic bug that must be caught by testing and review, not by trusting the kernel).

Lean can define grammars using \texttt{syntax} declarations and syntax categories. We implement a large subset of the Cat grammar; the supported subset is summarized in Section~\ref{sec:design:cat-subset}. This grammar layer allows Lean to recognize Cat tokens and build a \emph{concrete syntax tree} (CST), but it does not assign any semantics. To turn a CST into meaningful Lean definitions, we can use either macros or elaborator extensions.

In this project, we primarily use macros, because they provide a direct mapping from Cat surface syntax to Lean definitions. For example, Cat \texttt{let} bindings can be translated into Lean \texttt{def} declarations in a syntax-directed way. While an AST-based approach could offer more control, it would require more global bookkeeping (e.g., managing the environment and name resolution) and would significantly increase complexity.

We use \texttt{namespace}s to separate different memory models: each Cat model is elaborated into a corresponding Lean namespace. For example, the Cat \textsc{tso} fragment in Figure~\ref{fig:cat-tso} becomes a Lean namespace of the following form:
\begin{verbatim}
namespace TSO

def com_tso := rf ∪ co ∪ fr

def po1_tso := po ∩ (W * W ∪ R * M)

def ghb := po1_tso ∪ com_tso

def tso := Acyclic ghb

end TSO
\end{verbatim}

\subsection{Supported \texttt{cat} Subset (Targeting LKMM)}
\label{sec:design:cat-subset}
We show our work by supporting a subset of LKMM models, and the limitation is clear. The \texttt{cat} language is defined using an OCaml DSL, which gives it the flexibility to perform complex computations, such as defining procedures that compute relations. These computations often cannot be translated into logical propositions; in most cases, we cannot rely on arbitrary computations for formal proofs.

For example, LKMM defines a relation called \texttt{matched}, which is represented as follows:

\begin{verbatim}
(* Compute matching pairs of nested Rcu-lock and Rcu-unlock *)
let rcu-rscs = let rec
        unmatched-locks = Rcu-lock \ domain(matched)
    and unmatched-unlocks = Rcu-unlock \ range(matched)
    and unmatched = unmatched-locks | unmatched-unlocks
    and unmatched-po = [unmatched] ; po ; [unmatched]
    and unmatched-locks-to-unlocks =
        [unmatched-locks] ; po ; [unmatched-unlocks]
    and matched = matched | (unmatched-locks-to-unlocks \
        (unmatched-po ; unmatched-po))
    in matched
\end{verbatim}
This recursive function computes the RCU lock pairs. An approach is to encode the recursive function directly in Lean 4. However, Lean requires recursive definitions to be provably terminated. Structural recursion is automatically accepted, but for more general recursion, we must provide a termination proof (e.g., via a well-founded measure). In general, termination checking cannot be fully automated because determining whether arbitrary programs terminate is undecidable.

Another approach is to encode the recursion using a least fixed-point formulation. Since the data types in Cat are represented using sets and relations, the semantics can be translated into SMT formulas over relations and sets. In principle, the resulting constraints could be solved by an SMT solver. However, the SMT integrations available in Lean 4 are still relatively immature, making this approach unreliable in practice. Although the problem could also be formalized and proven directly in Lean, doing so would introduce significant additional complexity. Due to time constraints, we leave this aspect for future work.

\subsection{Design Choices}
The design involves several trade-offs arising from the limitations of Lean~4 and from aspects of the Cat language that are left implicit or undefined in the specification. The following subsections describe the key decisions made during implementation.

\textbf{Implicit Definitions}
Herd7 is a project under active development, so its documentation is sometimes outdated; it is necessary to consult the practical implementation directly. The tag names in Cat are converted into capitalized identifiers in the generated Lean code.

Besides the documentation, some aspects of the Cat language are implicit. For example, the manual presents the ``semantics of instructions'' as follows \citet{catManualSyntax11}:
\begin{quote}\small
\begin{verbatim}
instructions aevt [annotation 1, ..., annotation n]K X <j, f, rho, omega>,
  let typei : Ti = Jannotation iK X <j, f, rho, omega>, i in [1, n] in
  if forall i in [1, n]. typei = set /\
     forall type' : t in Ti. type' = tag /\
     forall e in evts-of(X). (kind-of(e) = aevt) ->
       (forall i in [1, n]. annot-of(e)i in Ti)
  then <j, f, rho, omega>
  else error.
\end{verbatim}
\end{quote}
This mainly specifies when a Cat program should return an error state, rather than giving a complete, executable meaning. As a result, we must make additional design choices to turn such fragments into a usable semantics. In addition to this, there are other implicit definitions, which we discuss in the definitions section.

\textbf{Reflexive-Transitive Closure Approximation}

The Cat language uses $\mathtt{e\text{*}}$ to denote the reflexive-transitive closure of a relation, and $\mathtt{e\text{+}}$ for the transitive closure. In a full model, these should produce every pair $(a, b)$ reachable by following zero or more (resp.\ one or more) edges, corresponding to \texttt{Relation.ReflTransGen} and \texttt{Relation.TransGen} in Lean~4. However, encoding these closures directly inside macro-generated definitions creates a practical difficulty: the resulting terms involve an existential quantification over arbitrarily long paths, which is not suitable for the kind of term-level definitional unfolding that tactics like \texttt{simp} and \texttt{aesop} rely on.

We instead compile $\mathtt{e\text{*}}$ to $r \cup \mathit{id}$ and $\mathtt{e\text{+}}$ to $r$, where $\mathit{id}$ is the identity relation. This means both operators produce a reflexive but not transitively closed result. In the specific contexts where these operators appear in LKMM---for example, $\mathtt{hb\text{*}}$ inside the propagation relation---the approximation is acceptable because the enclosing acyclicity check ensures that no cycle is inadvertently hidden. However, this is a known limitation: relations defined via $\mathtt{e\text{*}}$ may under-approximate the intended ordering in contexts that depend on full reachability.

\textbf{Primitive Relations and Ignored Directives}

The Cat standard library (\texttt{cos.cat}, \texttt{lock.cat}) defines several relations computationally, including the coherence order and the \texttt{fencerel} function, which computes the set of pairs $(a, b)$ such that a fence event $f$ exists with $(a, f) \in \mathit{po}$ and $(f, b) \in \mathit{po}$. Rather than implementing the full import pipeline, we treat \texttt{include} directives as no-ops and model the imported relations---as well as architecture-specific fence relations such as \texttt{mb} and \texttt{wmb}---directly as primitive fields of \texttt{CandidateExecution}. This is consistent with the view that the candidate execution is already the output of a lower-level analysis: herd7 computes these relations from the program state before invoking the Cat consistency check, so at the memory model level they can simply be treated as given inputs.

The Cat language also provides \texttt{flag} assertions, which are non-fatal warnings used to detect certain execution patterns (such as unmatched RCU lock/unlock pairs) without invalidating the execution. Since these assertions do not affect consistency, we discard them during elaboration. Concretely, the elaborator silently returns an empty command for both \texttt{include} and \texttt{flag} instructions.

\textbf{Program Order as ID Comparison}

In the Cat language, program order (\texttt{po}) is an abstract relation. We encode it concretely as $e_1.\mathit{id} < e_2.\mathit{id} \wedge e_1.\mathit{t\_id} = e_2.\mathit{t\_id}$, where $\mathit{id}$ is the unique event identifier and $\mathit{t\_id}$ is the thread identifier. This encoding assumes that event identifiers are assigned consistently with the program order within each thread, so that events appearing earlier in the program carry smaller IDs. It removes the need to maintain an explicit set of program-order edges, which simplifies the definition of \texttt{Events} and reduces the number of well-formedness conditions that litmus test authors must establish.

This design choice interacts with the \texttt{event\_id\_unique} axiom: since two events with the same ID are definitionally equal, the ID comparison faithfully represents distinctness, and the ordering it induces within a thread is consistent with the intended program order. Litmus test authors must therefore assign event IDs carefully, but in practice this is easy to satisfy by numbering events sequentially within each thread.

\textbf{From-Read as a Derived Relation}

In the Cat standard library, the from-read relation $\mathit{fr}$ is defined as $\mathit{rf}^{-1}; \mathit{co}$: a read $r$ is from-read of a write $w'$ if there is some write $w$ from which $r$ reads, and $w$ coherence-precedes $w'$. This captures the intuition that $r$ observed a write that was subsequently overwritten.

We follow this definition exactly, computing $\mathit{fr}$ as a derived field of \texttt{CandidateExecution} rather than a primitive one. Concretely, \texttt{fr'} is defined as \texttt{X.rf'.inv.comp X.co'}, which is transparent to the kernel: any goal involving \texttt{fr'} is unfolded automatically by \texttt{simp} into a goal about \texttt{rf'} and \texttt{co'}. This reduces the number of obligations that litmus test authors must discharge, since \texttt{fr'} does not need to be supplied manually for each test.

\textbf{Coercions between Set and SetRel}

Cat expressions freely mix sets of events and binary relations. For example, the identity-restriction operator $[\mathtt{S}]$ takes a set $S$ and produces the relation $\{(e_1, e_2) \mid e_1 = e_2 \wedge e_1 \in S\}$, while the domain of a relation is a set. To handle this fluently, we provide two coercions: \texttt{Set Event} can be coerced to \texttt{SetRel Event Event} via \texttt{SetRel.mkId}, and \texttt{SetRel Event Event} can be coerced to \texttt{Set Event} via \texttt{SetRel.dom}. Lean~4's coercion mechanism then inserts these conversions automatically wherever a set is used in a relation context or vice versa, so that the macro expansion of $[\mathtt{S}]$ can emit \texttt{SetRel.mkId s} without manually threading the coercion through every call site.

\subsection{Library Theorems and Proof Support}
\label{sec:design:theorems}
Cat relies on binary relations and set theory, including relational closures, irreflexivity, and related properties. Lean~4 supports two representations of binary relations: \texttt{Relation} (of type \texttt{Type → Type → Prop}) and \texttt{SetRel} (a set of pairs). Both representations are correct, but \texttt{SetRel} offers more flexibility for operations such as computing the domain and codomain. While Mathlib provides many relevant theorems, its coverage is insufficient for mechanizing memory models in full; consequently, we developed additional definitions and lemmas where needed.



