\chapter{Discussion}
\label{chapter:discussion}

In this chapter, we reflect on the design and results of LeanCat. We first discuss the limitations of the current implementation (Section~\ref{sec:limitations}), identifying the boundaries of the Cat subset we support and the challenges that remain open. We then compare LeanCat with related work (Section~\ref{sec:related-work}), situating it within the broader landscape of memory model mechanization and verification tools.

\section{Limitations}
\label{sec:limitations}

LeanCat achieves its stated goals, but several limitations remain. We discuss the most significant ones below.

\textbf{Partial Cat language support.} LeanCat supports a subset of the Cat language targeting SC, TSO, and LKMM. Several features are not yet supported: the RCU-related recursive constructs in LKMM (discussed in Section~\ref{sec:design:cat-subset}), the full \texttt{cross} operation from the Cat standard library, and architecture-specific Bell file features beyond the basic tag system. Extending support to these features would be necessary to handle more complex models such as RISC-V or Power.

\textbf{Incomplete BPF--LKMM comparison.} As discussed in Section~\ref{sec:bpf}, we were unable to prove fence-related ordering properties between BPF and LKMM, because no formal mapping between their respective fence primitives has been established in the literature. The partial result we prove --- that BPF is stronger than LKMM on the fence-free fragment --- relies on this gap and leaves the full comparison open.

\textbf{Parser limitations.} The Lean 4 parser cannot parse all Cat files as they are. Identifiers that begin with a single quote (\texttt{‘ONCE}, \texttt{‘RELEASE}, etc.) have no counterpart in Lean 4’s syntax and must be preprocessed into valid identifiers before parsing. Similarly, the hyphen character is not permitted in Lean 4 names, so Cat identifiers such as \texttt{read-once} must be rewritten using underscores. These preprocessing steps are handled automatically by LeanCat, but they add complexity to the elaboration pipeline.

\textbf{SMT solver integration.} An alternative approach to ITP-based verification is to encode Cat models into SMT formulas, as done by Dartagnan~\citep{10.1145/3776687}. SMT solvers can efficiently answer reachability questions for specific executions, but they do not produce proof certificates and their reasoning is difficult to inspect. An ITP is therefore more appropriate when the goal is to understand and verify the models themselves rather than to analyse specific programs. CVC5~\citep{barbosa2022cvc5} provides built-in support for binary relations and could reduce boilerplate by offloading certain obligations to the solver, but the SMT libraries available for Lean 4 are still in an early stage of development and cannot yet be relied upon for production proofs.


\section{Related Work}
\label{sec:related-work}

The ARM team mechanized the ARMv8 memory model in Coq~\citep{10.1145/3458926ArmedCat}, producing a formalization called RCoq that proves equivalences between different axiomatic formulations of the ARM model. Their mechanization is tightly coupled to the ARM architecture: the definitions and proofs are written by hand in Coq specifically for ARMv8. In contrast, LeanCat is architecture-independent --- it parses standard Cat specification files and introduces them as native Lean~4 definitions, so the same framework handles SC, TSO, MIPS, and LKMM without architecture-specific proof engineering.

\citet{DennisPhdThesis} develops Burrow, an Agda proof framework for verifying the correctness of binary translations between architectures (x86 to ARM, x86 to RISC-V). The proofs establish that a translation preserves all behaviors allowed under the source memory model, ensuring no new weak-memory behaviors appear on the target architecture. Burrow identifies common proof patterns across different architecture pairs and abstracts them into reusable combinators. However, Burrow targets a different problem than LeanCat: it verifies translation correctness rather than comparing memory models directly, and it does not reuse Cat files as specifications.

\citet{wickerson2017automatically} use the Alloy model finder, which is SAT-based, to automatically compare memory consistency models and discover distinguishing litmus tests. Given two models expressed in a Cat-like notation, their tool searches for executions that are allowed by one model but forbidden by the other, thereby checking whether one model is strictly weaker than another. This approach is highly automated and has been used to find subtle differences between models. However, the results are not machine-checkable proofs: when the tool reports that one model includes another, there is no proof certificate that can be independently verified. LeanCat trades some automation for this stronger guarantee --- our inclusion and acyclicity proofs are checked by the Lean~4 kernel.

\citet{kang2017promising} propose the Promising Semantics, an operational memory model designed to validate common compiler optimizations while ruling out out-of-thin-air reads. Unlike axiomatic models expressed in Cat, the Promising Semantics uses an operational style based on per-thread views and promises. The accompanying Coq mechanization verifies key metatheoretic properties of the model, including DRF theorems and compilation correctness to x86-TSO and Power. Because LeanCat operates within the Cat-based axiomatic framework, the two approaches are complementary: LeanCat can compare axiomatic models that serve as compilation targets, while the Promising Semantics addresses the design of source-level language models.

Dartagnan~\citep{10.1145/3776687} encodes Cat models into SMT formulas for bounded model checking of concurrent programs. As discussed in the previous section, this approach provides efficient automation but does not produce proof certificates. LeanCat can be seen as complementary: Dartagnan answers reachability questions about specific programs, while LeanCat proves general properties about the models themselves.

