\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

We’re entering an age of increasing hardware demands driven by AI training and inference needs, leading to a surge in specialized GPUs and accelerators. To achieve high performance, these architectures rely on complex, highly parallel execution across memory systems, where program behavior depends entirely on the underlying \emph{weak memory models} --- formal specifications that characterize which concurrent program outcomes are observable on a given architecture. To reason about such behaviors systematically, the Cat specification language represents program executions as graphs of binary relations, enabling concise axiomatic definitions of weak memory models. Many memory models---including those for RISC-V, ARM, x86, and the Linux Kernel Memory Model (LKMM)---are already defined in Cat. However, mechanizing these models in a proof assistant and constructing formal proofs about them remain largely orthogonal problems, with few existing solutions addressing both simultaneously. This gap becomes increasingly relevant as informal reasoning and ad hoc testing struggle to capture the full space of concurrent behaviors, and as precise relationships between models are needed for tasks such as compiler validation and cross-platform correctness.

In this thesis, we present LeanCat, a domain-specific proof framework for mechanizing weak memory models in Lean~4. LeanCat parses existing Cat specifications directly and imports their semantics into Lean~4 via its metaprogramming framework, eliminating the need to translate or rewrite memory model definitions. The framework provides the necessary primitives, abstractions, and library theorems to help users prove properties such as whether one memory model is weaker than another.

We implement a core subset of the Cat language targeting LKMM and demonstrate LeanCat through two case studies: checking litmus test executions (small concurrent programs that witness specific memory behaviors) and formally comparing memory models. In particular, we present a machine-checked proof that Sequential Consistency (SC) is strictly stronger than Total Store Order (TSO), reducing the comparison to a monotonicity argument over acyclicity constraints. Our results show that LeanCat significantly reduces the mechanization effort and boilerplate required for weak memory model proofs, and that the approach generalizes naturally to other model comparisons. Notably, these two case studies highlight complementary perspectives: litmus tests act as compact, targeted programs for exploring and distinguishing specific allowed or forbidden executions, whereas model comparisons establish global guarantees that hold across all programs.